Type Checker Generation using Reference Attribute Grammars

### 2024-12-09 presentationpics.tex
Commit final changes from way back

% disable spellcheck for this section
% LTeX: SETTINGS enabled=false
\lstdefinestyle{mystyle}{
  basicstyle=\ttfamily\footnotesize,
  breakatwhitespace=false,
  breaklines=true,
  captionpos=b,
  keepspaces=true,
  numbers=left,
  numbersep=5pt,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2
}
\lstset{style=mystyle}
\lstdefinelanguage{jrag}[]{java}{
  morekeywords={aspect, syn, eq, inh}
}
% LTeX: SETTINGS enabled=true

\setcounter{secnumdepth}{4} % Number subsubsections
%\setcounter{tocdepth}{4} % Include subsubsections in table of contents

% used to display the used files at the end. Select nofilelist as a package option to disable this
\listfiles % initialize

%\geometry{showframe}
%better like this?
\student{Nicholas Boyd Isacsson}{nicholas@isacsson.se}
%\students{Flavius Gruian}{Flavius.Gruian@cs.lth.se}{Camilla Lekebjer}{Camilla.Lekebjer@cs.lth.se}

\thesisnumber{LU-CS-EX: 2024-06} % Magic Number! Do not change unless Birger Swahn asks you to do so!
% default is Master. Uncomment the following for "kandidatarbete"/Bachelor's thesis
%\thesistype{Bachelor}{Kandidatarbete}

\newcommand{\CR}[1]{\textcolor{green!60!black}{[\textbf{CR}:#1]}}
%\newcommand{\CR}[1]{}

% For questions to people reviewing, remove before finalizing
\newcommand{\reviewquestion}[1]{\textcolor{red!80!black}{[\textbf{Review question}:#1]}}
%\newcommand{\reviewquestion}[1]{\textcolor{red}{\begin{itemize}\item Question: #1\end{itemize}}}
%\newcommand{\reviewquestion}[1]{}

%\title{Formatting a Master's Thesis}
\title{Type Checker Generation using Reference Attribute Grammars}
\svensktitel{Generering av Typkontrollerare med Referens-Attribut-Grammatik}

%\onelinetitle
%\twolinestitle
\threelinestitle
%\fourlinestitle

%\subtitle{A {\LaTeX} class}
%\company{The Corporation AB LTD Inc}
%\supervisors{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}{Don Jeer, \href{mailto:djeer@xy.lth.se}{\texttt{djeer@xy.lth.se}}}
\supervisor{Christoph Reichenbach, \href{mailto:christoph.reichenbach@cs.lth.se}{\texttt{christoph.reichenbach@cs.lth.se}}}
%\supervisor{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}
\examiner{Niklas Fors, \href{mailto:niklas.fors@cs.lth.se}{\texttt{niklas.fors@cs.lth.se}}}

%\date{\today}
\date{26th January, 2024}

\acknowledgements{
Sincere thanks to Christoph Reichenbach, whose enthusiasm for the field inspired me to pursue this project and whose guidance in matters of both implementation and writing was key to this thesis.
Additional thanks to Niklas Fors for providing several valuable insights, greatly simplifying certain components and Alexandru Dura for providing some guidance to JastAdd internals.

I would also like to thank my partner Lo-Isabel Krantz Andrée as well as my family, friends, and Moritz for their support throughout the duration of this project.

%If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.
%
%We would like to thank Lennart Andersson for his feedback on this template.
%
%We would also like thank Camilla Lekebjer for her contribution on this template, as well as Magnus Hultin for his popular science summary class and example document.
%
%Thanks also go to the following (former) students for helping with feedback and suggestions on this template: Mikael Persson, Christoffer Lundgren, Mahmoud Nasser.
\vspace*{\fill}
\hspace*{\fill}
\includegraphics[scale=0.5]{assets/cat.pdf}
}

\theabstract{
%Your abstract should capture, in English, the whole thesis with focus on the problem and solution in 150 words. It should be placed on a separate right-hand page, with an additional \textit{1cm} margin on both left and right. Avoid acronyms, footnotes, and references in the abstract if possible.
%Leave a \textit{2cm} vertical space after the abstract and provide a few keywords relevant for your report. Use five to six words, of which at most two should be from the title.

  Static type checkers are a popular method for early detection of bugs in computer programs, by ensuring the correct data types are used throughout the program.
  % RAGs may not be the most convenient format for typing rules
  One approach for implementing type checkers are Reference Attribute Grammars (RAGs), a formalism for specifying the static semantics of programming languages by declaring attributes on an abstract tree-representation of the program, and is used by the JastAdd meta-compilation system.

  This report contains a study on the usability of JastAdd to programmatically generate executable type checkers from a subset of possible typing rules, known as 'syntax-directed' typing rules.
  This aims to provide integration with existing compilers implemented with JastAdd, as well as a playground for type systems usable as an educational tool.

  % 'Syntax-directed' needs explaining if used, "a subset known as syntax-directed"
  We find RAGs to be a flexible model for syntax-directed typing rules, and our software was able to support small languages from research literature.
  It proved difficult to provide compile-time verification of the typing rules, and further work is required to extend the compiler to support type environments, a necessity for most real-world type systems.
}

%\keywords{MSc, BSc, template, report, style, structure}
\keywords{type checking, type theory, code generation, reference attribute grammars}

%% Only used to display font sizes
\makeatletter
\newcommand\thefontsize[1]{{#1 \f@size pt\par}}
\makeatother
%%%%%%%%%%

\begin{document}
\renewcommand{\bibname}{References}

\makefrontmatter
\chapter{Introduction} % should give the background of your work. Important parts to cover:
% Give the context of your work, have a short introduction to the area.
While writing software, developers often make mistakes.
Catching these mistakes, as early as possible, can prevent these bugs from ever reaching production code.
%\CR{Possible papers to cite as evidence:
%  \url{https://dl.acm.org/doi/abs/10.1145/1985793.1985889}
%  \url{https://www.usenix.org/legacy/event/sec04/tech/full_papers/johnson/johnson_html/}
%  \url{https://ieeexplore.ieee.org/abstract/document/10190394}}
%\CR{These papers add static checkers on top of existing languages to find additional bugs; you can also
%  mention statically typed languages, but those generally don't require frequent changes to the type checker implementation, so they're less
%  compelling to motivate your work.}
Type checkers are one of the most common static checking methods, ensuring that the correct data types are used.
While some languages include type checkers in their compilers, work has also been done to add static type checkers to existing languages.
This requires writing large amounts of specialised code on a per language basis, but has shown to be effective at detecting bugs\cite{10.1145/1985793.1985889}\cite{10.5555/1251375.1251384}\cite{10190394}.
Furthermore, making changes to the typing rules of a language can require extensive rewrites in many different parts of the type checker's codebase.

% Define the problem you are solving (or trying to solve).
%\CR{Prior work has already looked into pluggable type checkers.
%  Why is translating them into JastAdd interesting?
%  (aside from theoretical questions, you can also use
%  pragmatic reasons, such as integration into our courses.)}
In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of Reference Attribute Grammars (RAGs).

Pacak et al. have previously implemented a compiler for generating incremental type checkers using declarative logic programming language Datalog\cite{Pacak}.
It focused on defining a way of expressing typing rules as a series of relations solvable by the Datalog compiler.
Our use of RAGs allows us instead to build our analysis directly onto a representation of the program's structure.
JastAdd is also an interesting target because it has already been used to implement compilers of several languages, to which our typing rule implementations could potentially be integrated.
It is also used extensively in courses within the Department of Computer Science at LTH, where our project could serve as an educational tool, providing students an interactive way to define and test simple type systems.

Typing rules form a complex, often ambiguous, and infinite relation.
To be able to translate them into JastAdd code, we will need to define a limited subset of syntax-directed typing relations.
We identify a number of different kinds of typing rules, used in previous work in the field\cite{Pierce}, which share a common set of features for us to implement.
We then devise a strategy to translate these rules into snippets of iterative Java code, which can be evaluated by JastAdd.

%\CR{Make clear to the reader what the challenges are:
%      How are these two formalisms similar, how do they differ (at a high level)?
%      You can use a simple typing rule as an example if it helps (very optional).
%      Hint: look at determinism in evaluation (syntax-directedness), unification, but stick to intuition in this chapter; you can elaborate in Chapter 2 or later if you think it helps.}

We define a custom parsable language for typing rules adapted from natural semantics, and introduce a strategy to compile it into RAG attributes.
Our case studies show that simple type systems can be systematically compiled into attributes on the Abstract Syntax Tree (AST), and the tree syntax provides a natural approach for recursive evaluation.
Overall, this initial exploration shows that RAGs are a flexible approach for implementing type checkers for syntax-directed typing rules, though there are difficulties when implementing rule verification at compiler runtime.
%\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
%  We evaluate the approach against Foo, and demonstrate Bar.
%  Our case studies show Quux.
%  Overall, this initial exploration shows \ldots''}

% Specify your contributions. What does this particular work/report bring to the research are or to the body of knowledge?
\section{Research questions}
To formulate our goals more concretely, we have set out to answer the following research questions:
%\CR{Give the RQs numbers so you can refer to them later.  Common form: RQ1, RQ2 etc.; some authors use RQ2.3 if they split RQ2 into multiple subquestions (this can happen later).}
\newcommand{\rqone}{RQ1: What kinds of typing rules can we translate using RAGs?}
\newcommand{\rqtwo}{RQ2: How can we implement the translations and what algorithms should we use for the output code?}
\newcommand{\rqthree}{RQ3: What are the challenges in translating typing rules into JastAdd-based RAGs?}
\begin{itemize}
  \item \rqone
    %\CR{``kinds'' of typing rules: right now the reader isn't aware that there even are different kinds (see my earlier comment on ``what the challenges are''-- sketching this out should suffice.)}
  \item \rqtwo
    %\CR{What does ``form'' mean, here?  How broad is ``implement''-- are you talking algorithms (which I think you are) or including technical challenges (which I think you have in the next RQ)?}
  \item \rqthree
    %\CR{I'd go with `challenges in translating' instead of `potential issues of translating', language-wise}
\end{itemize}

\section{Chapter outline}
%\CR{%paper $\to$ thesis; also,
 % this always comes at the very end of the introduction}
The rest of this thesis is outlined as follows.
% This chapter introduces the
Chapter 2 presents the necessary background information for the report.
Chapter 3 describes the design and implementation of our program, followed by demonstrations of its results in chapter 4.
Chapter 5 discusses the results in a more general sense, before the conclusion in chapter 6.

\chapter{Background}
%\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
%\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinking ``I know why I want to read this''}
This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
Section \ref{abstractsyntaxtrees} explains abstract syntax trees, a simplified tree-based representation of a program, suitable for performing static analysis on.
These are extensively utilised in almost every stage of our project, both in terms of our implementation and the input to the program.
%\CR{Consider saying just a little more about them here, for readers who have no idea what ASTs are and why they matter for
%    analysing programs.}
Section \ref{typetheory} details type systems, by explaining each of the component pieces, followed by a description of their use within static type checking.

\section{Abstract syntax trees}\label{abstractsyntaxtrees}
\begin{figure}[]
  \centering
  \begin{tikzpicture}[
      every node/.style = {align=center, line width=0.8pt},
      every path/.style = {draw, -latex},
      astnode/.style={draw=black},
      %greennode/.style={shape=circle, draw=green, line width=2},
      %rednode/.style={shape=circle, draw=red, line width=2}
    ]
    \node (if)   [astnode]                             {If};
    \node (then) [astnode, below=of if]                {True};
    \node (else) [astnode, right=of then]              {False};
    \node (cond) [astnode, left=of then]               {GreaterThan};
    \node (one)  [astnode, below = of cond.south west] {Int\\ \textcolor{gray}{value: 1}};
    \node (zero) [astnode, below = of cond.south east] {Int\\ \textcolor{gray}{value: 0}};

    %\node () [above right = 0.3cm and -1cm of cond] {cond};
    %\node () [above left = 0.1cm and -0.7cm of then] {then};
    %\node () [above left= 0.3cm and -0.5cm of else] {else};
    %\node () [above= 0.3cm of one] {left};
    %\node () [above= 0.3cm of zero] {right};

    \draw (if.south)   to (cond.north);
    \draw (if.south)   to (then.north);
    \draw (if.south)   to (else.north);
    \draw (cond.south) to (one.north);
    \draw (cond.south) to (zero.north);
    %\draw (if.south)   to [edge label=Cond]  (cond.north);
    %\draw (if.south)   to [edge label=Then]  (then.north);
    %\draw (if.south)   to [edge label=Else]  (else.north);
    %\draw (cond.south) to [edge label=Left]  (one.north);
    %\draw (cond.south) to [edge label=Right] (zero.north);

  \end{tikzpicture}
  \caption{Example of an abstract syntax tree}
  \label{astexample}
\end{figure}
The syntax of a language is the definition of the symbols and structures which make up a valid program.
%\CR{What's a syntax?}
An abstract syntax tree (AST) is a simplified tree-based representation of the syntax.
Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
It is abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this semantic information.

This makes it easier to write analysis based on the behaviour of a program, without concerning oneself with its exact formulation.
Several different formulations of a program with identical behaviour may be abstracted into identical ASTs.
The hierarchical structure of the tree lends itself especially well for recursive analyses, and eases the passing of information between connected structures by connecting them as parent and children nodes.
%\CR{what is ``this information'', concretely?}
Figure \ref{astexample} shows an example of an abstract syntax tree representation of the expression \verb|if 1 > 0 then true else false|.
%A statement like \verb|if (true) then {1} else {0}| may instead be represented by an if node, with named branches pointing towards the condition and the two values.
%\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
%\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}

%\CR{Consider giving the syntax of a simple expression language in BNF here, since you're already discussing ``Type syntax'' later.}
%\begin{figure}[h]
%\begin{lstlisting}[]
%Expr   -> Expr "+" Term | Term
%Term   -> Term "*" Factor | Factor
%Factor -> Int | "(" Expr ")"
%\end{lstlisting}
%  \caption{\reviewquestion{One like this? I'm unsure how to tie it in effectively.}}
%  \label{typesyntaxspec}
%\end{figure}

\section{Type theory}\label{typetheory}
In this section we will discuss the elements of type theory required to understand this thesis.
Firstly, we will discuss what types are and give some examples of their use.
Next, we will describe type systems, which form a relation between types and expressions.

\subsection{Values and expressions}
The main purpose of most computer software is to handle and manipulate data.
Each of these individual pieces of data is a value, which could be anything from the number 4 or the text `Hello!'.

Expressions are a broader term, indicating any syntactic structure which evaluates to a value.
This includes but is not limited to values, which evaluate to themselves, but also constructs of values, operators or function calls, etc.

\subsection{Data types}
%\CR{think about the connection between types and values here.  You may even want to talk about values in a separate subsection.}
Data types can be considered a categorisation of values which share common properties, such as supporting numerical operations or accessing a certain element in a list of characters.
While expressions may be arbitrarily complex, nested structures, by assigning it a type, we can perform useful analysis on it, without having to evaluate its value.
%\CR{One mental image that helped me a lot to understand types is to see values  on one side (let's say on the left) and their types on the other (let's say the right).
%  For each concrete (and possibly very complex) operator on the left-hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
%  This symmetry between ``concrete'' and ``abstract'' only works if you make values part of your cast of characters for this discussion.
%  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractions over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
%}

%several different types of data in one program, such as numbers or text.
Most programming languages come with a number of built-in data types, from different kinds of numbers (integers, floating point) to strings or lists and arrays.

%\CR{if you discuss values, you can use the intuition of ``grouping values'' by certain properties (such as supported operations) to make your point clearer.}
Different parts of programs may have different constraints on the types of expressions, such as a print function requiring a string of text, or a plus function requiring some kind of number.
%\CR{Very good point about types of expressions, but again, this is likely easier to understand by linking expressions and values.
%  I strongly recommend using a definition of the form ``all values are expressions, but not all expressions are values''.}

\subsection{Type checking}
Verifying that these constraints are upheld can assist us in catching programming mistakes, preventing the wrong data types from being used in the wrong places, which may have led to erroneous or unpredictable behaviour.
There are two main approaches to ensuring the constraints are upheld, dynamic and static type checking.
As our project generates a static type checker, we will focus our explanations on that topic, but provide a quick overview of their differences here.
%\CR{Why would we even want to ensure that they are upheld?}

Dynamic checking is performed at runtime, with each type checked only when needed.
While this means no checks are performed unnecessarily, it makes it harder to verify the constraints over the whole program, as it will not be tested until each specific part is executed.

Static checking is performed on the code before it is run, usually as part of compilation, preventing many faulty programs from being run at all.
However, since it lacks knowledge of what happens at runtime, it is required to take a more conservative approach, flagging potential errors even in code which will never executed.
Static checkers analyse the structure of the source code, often in the form of an abstract syntax tree.

%\CR{Why would I want static checking?  Dynamic only checks when needed, isn't that more efficient?}

\subsection{Type systems}
A type system typically consists of two parts, a type syntax, describing the available types of a language; and typing rules, denoting how types should be assigned to expressions within the language.
%\CR{Why do the types need a syntax?
%    (There are other approaches to types; for your own sanity, I recommend deferring to Pierce here and calling this ``typical approaches'')}

\subsubsection{Type syntax}\label{typesyntax}
We're using the term `type syntax' to mean a definition of the valid types within a language, often referred to simply as the set of types.
%\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
%\reviewquestion{I believe Pierce simply refers to it as the "set of types". I seem to have got the term type syntax from EDAP15 \url{https://fileadmin.cs.lth.se/cs/Education/EDAP15/2022/web/skills/type-system.html}. In my opinion it works better to have a clear name for it, so I'm keeping it for now, unless you have further thoughts.}
It can consist of fundamental types, such as integers or booleans.
There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
This enables differentiating lists that store integers from those that store strings and ensuring that any item added to a list must be of the same type as the type parameter.

Figure \ref{typesyntaxspec} shows an example specification of a type syntax, consisting of only two types, integers and booleans.
This specification is used for all the languages used in our project.
%\CR{A bit vague; consider giving an example.}
\begin{figure}[]
\begin{lstlisting}[]
abstract Type;

Bool : Type;
Int  : Type;
\end{lstlisting}
  \caption{Type syntax consisting of two types, for boolean and integer values, written as a JastAdd AST specification (see Section \ref{astdef}).}
  \label{typesyntaxspec}
\end{figure}

\subsubsection{Typing rules}
The typing relation, notated \lstinline{t : T}, is the link between the expressions of a language and the types of the type syntax.
%\CR{It is \emph{the} link, even.}
The left-hand side is an expression in the object language, typically written in its native syntax, the right-hand side a type from the aforementioned type syntax.

Typing rules are a manner of inference rules, which in their simplest form, consist of no more than a typing relation.
The typing rule \lstinline{T-True} seen in Figure \ref{typingruleexampletrue}, for example, simply assigns the type of \lstinline{Bool} to the expression \lstinline{true}.
Additionally, the typing rules in these examples have a name, given in parentheses next to the rule, though this is merely for reference.

\begin{figure}[h]
  \begin{equation}
    \tag{T-True}
    \text{true} : \text{Bool}
  \end{equation}
  \caption{Typing rule for a true value}
  \label{typingruleexampletrue}
\end{figure}

Typing rules may also have premises, a number of conditions that must hold for the typing rule to apply.
%\CR{\emph{that} must hold (``which'' relates a set to a superset: ``carrots, which are vegetables'', while ``that'' restricts a set to a subset: ``vegetables that are carrots'')}
The notational convention here is to put the premises above a bar, with the conclusion underneath it.

\begin{figure}[h]
  \begin{equation}
    \tag{T-If}
    \frac{
      t_1 : \text{Bool}\quad
      t_2 : \textsf{T}\quad
      t_3 : \textsf{T}
    }{
      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \textsf{T}
    }
  \end{equation}
  \caption{Typing rule for an if-expression}
  \label{typingruleexampleif}
\end{figure}

\newcommand{\ty}[1]{\textcolor{blue!20!black}{\textsf{#1}}}
\newcommand{\tyvar}[1]{\textsf{#1}}
Figure \ref{typingruleexampleif} shows an example rule, typing an if-expression.
The bottom line, the conclusion, contains the keywords of the if-expression and three free variables $t_i$ which represent any term within the language.
Above the bar are three premises, placing constraints on the three variables, which must hold for the typing rule to apply.
The first constrains $t_1$ to the type bool, the second two constrain $t_2$ and $t_3$ to a type variable \tyvar{T}, which represents any single arbitrary type within the type syntax.
While \tyvar{T} may be any type, it must consistently be the same type within any single evaluation of the rule, i.e. the two terms must have the same type.
This \tyvar{T} also appears in the conclusion, as the resulting type of the if-expression, meaning the result of the entire expression will be that of the $t_2$ and $t_3$ terms.
%\CR{Consider typesetting T differently, e.g. as \tyvar{T}}
%\CR{\tyvar{T}, $t_1$, $t_2$, $t_3$ are \emph{free} variables here, logically speaking}.

\subsubsection{Evaluating typing rules}
While performing static type checking, the typing rules form the basis of our analysis.
By analysing the program, we attempt to match the syntactic form with the typing rules, to find the type of the expression.
If no matching typing rules can be found, a type error is produced.

The simplest program possible with the typing rules T-True and T-If, consists of the single value \lstinline{true}.
It can immediately be matched to the rule T-True, giving the type \lstinline{Bool}.

A somewhat more complex case can be expressed as \lstinline{if true then true else }\\\lstinline{true}.
The only rule that matches the structure, is T-If, with \lstinline{true} values taking the place of each of the $t_i$ variables.
To evaluate this rule, we are thus required to recursively evaluate the types each of our $t_i$ variables, to ensure they match the premises.
In this case, the type of $t_1$ evaluates to \lstinline{Bool}, matching the first premise of T-If.
The types of both $t_2$ and $t_3$, represented as the type variable \tyvar{T}, also evaluate to \lstinline{Bool}.
Since both the uses of this type variable within the premises are consistent, all three of the rule's premises have been fulfilled, and the entire expression can be typed as \tyvar{T}, which has been found to correspond to \lstinline{Bool}.

\begin{figure}[h]
  \begin{equation}
    \tag{T-Zero}
    \text{zero} : \text{Int}
  \end{equation}
  \caption{Typing rule for a zero value}
  \label{typingruleexamplezero}
\end{figure}

Adding a second value of a different type allows us to express incorrectly typed programs, for example the program \lstinline{if zero then zero else true}.
Recursively evaluating this according to T-If, we find that the premise concerning $t_1$ no longer upholds, as the type of it now evaluates to \lstinline{Int}.
Additionally, the type variable \tyvar{T} typed by both $t_2$ and $t_3$ now has contradictory use, being assigned both as the type \lstinline{Int} and \lstinline{Bool}.
As this program cannot be found to satisfy any set of typing rules in our type system, it is found to be invalid.

%\CR{Please explain the purpose of typing rules in static type checking.  Using just your two examples you would be able to show a well-typed program, and if you add one more rule (e.g. for integers), you can show a type error.  No need to build the full derivation tree, but it's important to mention recursion here.  This entire bit may be worth a separate subsubsection.}

\section{Reference Attribute Grammars}
This section introduces the subject of Reference Attribute Grammars (RAGs), first explaining the theory behind them, then introducing JastAdd, an implementation of RAGs within a Java meta-compilation system\cite{JastAdd}.

\subsection{Reference Attribute Grammars}
Reference Attribute Grammars are an extension to Attribute Grammars (AGs), a formalism used for specifying the static semantics of programming languages.
%\CR{Generally only for the static semantics (RAGs can also do dynamic semantics, if you allow circular attributes, but that's not worth mentioning here.)
%Please also connect ``semantics'' to ``type checking'' (via ``static semantics'')}
AGs provide a systematic way to define computations on the nodes of an abstract syntax tree, associating attributes with language constructs and are commonly used in the context of compiler construction, where they help define the translation of source code into executable code\cite{AGs}.

RAGs extend these properties by introducing a new kind of attribute, allowing references to other nodes in the AST, and through the references accessing the attributes of other nodes.
This provides a simplified and more flexible method for interconnectedness between nodes\cite{RAG}.

%\subsection{Non-Terminal Attributes}
%Our project also uses not\cite{higherorderattributegrammars}
%\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are not a necessary feature for RAGs.
%Please introduce them in a separate section: you can very directly connect them to the idea of ``type syntax''.}

\subsection{JastAdd}
JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion\cite{JastAdd}.
Our project uses it both for the implementation of our compiler and as a compilation target, as it outputs JastAdd code.
%\CR{It's worth telling the reader here why you are explaining some details about JastAdd.}

\subsubsection{AST specification}\label{astdef}
\begin{figure}[]
\begin{lstlisting}[]
Expression ::= Term;

abstract Term;
True  : Term;
False : Term;
Or    : Term ::= Left:Term Right:Term;
Zero  : Term;

\end{lstlisting}
  \caption{AST specification for the Bools language.}
  \label{astspecexample}
\end{figure}
One of the key components of JastAdd projects are the \lstinline{.ast} files, which declare the structure of the abstract syntax tree.
This specification is then used by JastAdd to generate Java classes representing the nodes of the AST.
These AST classes can have a variety of forms, supporting both abstract classes and subclasses.
They can also contain components such as Java types, tokens and child nodes, and these components may also be optional or list components.

Figure \ref{astspecexample} shows a small example of the specificaton syntax.
Each of the node definitions begins with a class name.
\lstinline{True : Term;} indicates that the \lstinline{True} node is a subclass of the abstract class \lstinline{Term}.
If a node has child nodes, these are provided after the \lstinline{::=}, and may either have a name and a type separated by a colon, or only a type.
For a detailed overview of the syntax, see the reference manual\cite{jastaddref}.

\subsubsection{RAG Modules}
\begin{figure}[]
\begin{lstlisting}[language=jrag]
aspect TypingRules {
  syn Type Zero.type() {
    return new Int();
  }
}
\end{lstlisting}
  \caption{Snippet of a .jrag file, declaring a type attribute for the Zero AST node}
  \label{jragexample}
\end{figure}
The RAG modules are written in \lstinline{.jrag} files and declare the attributes to be included in the classes generated by the AST specification.
Figure \ref{jragexample} shows a minimal example.
The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.

\chapter{Approach} % should contain a description of your solution(s), with all the theoretical background needed. On occasion this is replaced by a subset or all of the following:

%\CR{Missing: introduction to this chapter'}
We begin this chapter with a section presenting our language for writing typing rules, an adaptation of the natural semantics used to notate typing rules in the academic context.
In Section \ref{sectionarchitecture} we proceed with presenting the overall architecture of our software, including an overview graph in Figure \ref{overviewgraph}, which may be useful to bear in mind while reading the rest of the report.
Section \ref{sectionimplementation} goes into the details of our implementation, how we generate code for different parts of the typing rules.

\section{Language design}\label{sectionlanguage}
\begin{figure}[]
\begin{lstlisting}[]
ruleset = rule_list;

rule_list =
    rule
  | rule_list rule;

rule =
    '[' uppercase_id ']' premises_list '---+' formula
  | '[' uppercase_id ']' formula;

premises_list =
    formula
  | premises_list ',' formula;

formula = term ':' tyterm;

term =
    uppercase_id '(' term_list ')'
  | uppercase_id '(' ')'
  | uppercase_id
  | lowercase_id;

tyterm =
    uppercase_id
  | lowercase_id;

term_list =
    term
  | term_list ',' term;

uppercase_id = "[A-Z][A-Za-z0-9-]";
uppercase_id = "[a-z][A-Za-z0-9-]";
\end{lstlisting}
  \caption{A definition of the concrete syntax for the typing rule language in a notation similar to Backus-Naur form (BNF).}
  \label{trconcretesyntaxspec}
\end{figure}
%\CR{After reading, I would now say that it's preferable to me if you keep the language definition separate
%  from its implementation.  You can use Figure 3.4 as the abstract syntax, to complement the current discussion.
%  This would be before the current 3.1, then.
%  An advantage of this is that you can then start your ``Architecture'' discussion (in 3.2) by mentioning
%  the complex dependencies, i.e., that processing a type system specification requires you to know the abstract syntax
%  of the object language as well as of the types.}

\begin{figure}[]
\centering
\begin{minipage}{.45\textwidth}
  \centering
  \begin{equation}
    \tag{T-True}
    \text{true} : \text{Bool}
  \end{equation}
  %\caption*{Pierce syntax}
  %\label{fig:sub1}
\end{minipage}%
\hspace{.1\textwidth}\begin{minipage}{.45\textwidth}
  \centering
\begin{lstlisting}[]
[T-True]
True : Bool
\end{lstlisting}
  %\caption*{ASCII representation}
  %\label{fig:sub2}
\end{minipage}
\caption{A typing rule for the value true in natural semantic syntax on the left and its equivalent ASCII representation on the right}
\label{trexampletrue}
\end{figure}
\begin{figure}[]
\centering
\begin{minipage}{.45\textwidth}
  \centering
  \begin{equation}
    \tag{T-If}
    \frac{
      t_1 : \text{Bool}\quad
      t_2 : \text{T}\quad
      t_3 : \text{T}
    }{
      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \text{T}
    }
  \end{equation}
  %\caption*{Pierce syntax}
  %\label{fig:sub1}
\end{minipage}%
\hspace{.1\textwidth}\begin{minipage}{.45\textwidth}
  \centering
\begin{lstlisting}[]
[T-If]
t1 : Bool,
t2 : t,
t3 : t
-----------------
If(t1, t2, t3): t
\end{lstlisting}
  %\caption*{ASCII representation}
  %\label{fig:sub2}
\end{minipage}
\caption{A typing rule describing an if-expression in natural semantic syntax and its equivalent ASCII representation}
\label{trexampleif}
\end{figure}

%- Figure of example natural deduction rules (from Pierce?) and ASCII representation
%No infix notation
%No environments

To express typing rules, we have defined an ASCII representation of the natural semantics syntax used by Pierce\cite{Pierce}, with its concrete syntax given in Figure \ref{trconcretesyntaxspec}.
Our language can express a subset of the notation used by Pierce, with some differences.

The largest difference lies in the notation used for the expression to be typed.
Pierce uses the object language's native syntactic form, whereas our representation is based on the corresponding AST nodes.
Node names are consistent with those used by JastAdd, followed by parentheses containing variable names to denote their child nodes in the order of their definition.
This change is necessary to allow us to type arbitrary languages without needing to understand their syntax and grammar, though it does make translating rules from one syntax to the other less straightforward.

>>>>

\lstdefinestyle{mystyle}{
  basicstyle=\ttfamily\footnotesize,
  breakatwhitespace=false,
  breaklines=true,
  captionpos=b,
  keepspaces=true,
  numbers=left,
  numbersep=5pt,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2
}
\lstset{style=mystyle}
\lstdefinelanguage{jrag}[]{java}{
  morekeywords={aspect, syn, eq, inh}
}
\begin{document}
\begin{equation}
  \tag{T-If}
  \frac{
    t_1 : \text{Bool}\quad
    t_2 : \text{T}\quad
    t_3 : \text{T}
  }{
    \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \text{T}
  }
\end{equation}

  \resizebox{\textwidth}{!}{%
  \begin{tikzpicture}[
      every node/.style = {align=center, line width=0.8pt},
      every path/.style = {draw, -latex},
      artifact/.style={draw=black},
      userinput/.style={draw=red},
      generated/.style={draw=blue},
      process/.style={ellipse, draw=black}
      %greennode/.style={shape=circle, draw=green, line width=2},
      %rednode/.style={shape=circle, draw=red, line width=2}
    ]
    \node (typerules) [userinput]                    {Typing rules\\definition };
    \node (trparsing) [right=of typerules, process] {Scan,\\parse,\\build};
    \node (trparser)  [above=of trparsing, artifact]          {Typing rule parser};
    \node (trast)     [right=of trparsing, generated]          {Typing rules AST};

    %\node (typecheckgen) [below=of trast, process, text width=5cm] {
    %  \textbf{Type checker generation}\\
    %  \hrule
    %  \begin{itemize}[noitemsep,topsep=2pt]
    %    \item Copy template
    %    \item Generate type checker from typing rules
    %  \end{itemize}
    %};
    \node (typecheckgen)       [below=of trast, process] {Generate\\type checker};
    \node (typingrulecompiler) [right=of trast, artifact]          {Typing rule\\compiler };
    \node (typechecker)        [below=of typecheckgen, generated]   {Type checker };

    \node (olast)     [below=of typerules, userinput] {AST specification};

    \node (outputtypechecking) [below=of typechecker, process] {Perform\\typechecking};
    \node (inputast)           [left=of outputtypechecking, generated]    {Input program\\AST};
    \node (outputparsing)      [left=of inputast, process]     {Scan,\\parse,\\build};
    \node (typeerrors)         [right=of outputtypechecking, generated]   {Type errors};
    \node (inputprog)          [left=of outputparsing, artifact]         {Input program};

    %\node (template) [below=of inputast, text width=6cm] {
    %  \textbf{Type checker template}\\
    %  \hrule
    %  \begin{itemize}[noitemsep,topsep=2pt]
    %    \item Scanner, parser, builder
    %    \item Project files
    %    \item Type checking algorithm 
    %  \end{itemize}
    %};
    \node (templateparser)        [below=of outputparsing, artifact]                               {Tree-based\\parser};
    \node (templateprojectfiles)  at (inputast |- templateparser)           [artifact]           {Project files};
    \node (typecheckingalgorithm) at (outputtypechecking |- templateparser) [artifact]           {Type checking\\aspect };
    \node (templatetypesyntax)    [right=of typecheckingalgorithm, artifact]                       {Type syntax\\AST };
    \node (template)              [dashed, fit=(templateparser) (typecheckingalgorithm) (templatetypesyntax), artifact] {};
    \node                         [above, draw=none] at (template.north)                 {Support library};
